Tail associations and their impact on extinction risk

### 2019-10-30 MainText.Rmd
edited through methods

<!--FIG : pedagogical fig showing 4 types of tail associations: 
extreme left, moderate left, moderate right, extreme right-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=15 cm]{./Results/introfig_copula.pdf}
\caption{Two stochastic quantities can have same Spearman 
correlation, but different degrees of tail association: 
(A) extreme left-tail association, (B) moderate left-tail association, (C) 
moderate right-tail association, and (D) extreme right-tail association.
In all cases, Spearman's $\rho$ was the same, $\rho = 0.875$,
up to sample variation (the sample correlation for each dataset is displayed
on its panel). The extreme left-tail (respectively, right-tail) dependent 
case is perfectly correlated below (respectively, above) the threshold of 0.
See Methods for information on how bivariate noise was generated with
asymmetric tail associations.
\label{fig_intro_cop}}
\end{center}
\end{figure*}

@@

The stochastic quantities $s_e(t)$ and $\tilde{f}_a(t)$ were generated 
to have the same overall association (Spearman's $\rho$ equal to $0.875$,
up to sampling variation), but different degrees and types of 
asymmetry of tail association in different simulations, as follows.
For a simulation of length $T$, we used methods that can 
generate $T$ independent pairs $(u(t),v(t))$ for $t=1,\ldots,T$ with
$\rho=0.875$ (up to sample variation), but with either extreme left-,
moderate left-, moderate right-, or extreme right-tail association 
(e.g., Fig. \ref{fig_intro_cop}). The methods we used generated
$(u(t),v(t))$ such that $u(t)$ was uniformly distributed on $(0,1)$,
and likewise for $v(t)$. We then applied a monotonic transformation
to the $u(t)$ so that the resulting numbers were distributed according to
a beta distribution with parameters
$\alpha=3$ and $\beta=2$, 
and we used those values for the $s_e(t)$. We also applied a (different)
monotonic transformation to the $v(t)$ so that the resulting numbers were
distributed according to a gamma distribution with parameters $k=2$
and $\theta=1$, and we used those values for the $\tilde{f}_a(t)$. Because
the $u(t)$ and $v(t)$ were uniformly distributed, the inverse 
cummulative distribution functions of the desired beta and
gamma distributions were the transformations. Because these
were monotonic transformations, the Spearman correlation of 
$s_e(t)$ and $\tilde{f}_a(t)$ was the same as that of 
$u(t)$ and $v(t)$. The quantities $(u(t),v(t))$ for $t=1,\ldots,T$ were
generated as described in Appendix \ref{SM-makenoise}.

<!--calculation of ext risk with age-structured model-->
To calculate extinction risk for a given $\Omega$ and choice of tail dependence, 
we simulated the model for $25$ time steps, $10000$ times,
each simulation starting from $E(t)=5$ and $A(t)=10$. The population for a simulation
was considered to have gone extinct if the total population $E(t)+A(t)$ ever 
went below an extinction threshold (we used a threshold of $2$) within the 
$25$-time-step time horizon. Extinction risk was the fraction of $10000$
simulations that went extinct.

The variables $u(t)$ and $v(t)$ are visibly asymmetrically tail associated (Fig. \ref{fig_intro_cop}),
but we have instead used the transformed variables $s_e(t)$ and $\tilde{f}_a(t)$
in our model. In what sense are $s_e(t)$ and $\tilde{f}_a(t)$ asymmetrically
tail associated? Given two boundary quantities $l_b$ and $u_b$ such
that $0 \leq l_b < u_b \leq 1$, and letting $x(t)$ be the rank of 
$u(t)$ in the set $\{u(t) | t=1,\ldots,T\}$ and letting 
$y(t)$ be the rank of 
$v(t)$ in the set $\{v(t) | i=1,\ldots,T\}$ (smallest elements
are here defined to have rank 1),
consider the portions of the 
lines $x+y = 2l_b$ and $x+y=2u_b$ which meet with the unit square
(Fig. \ref{PartialSpearmanFig}). 
<!--***DAN: need to make this figure-->
We can then define the *partial Spearman correlation*,
\begin{equation}
\cor_{l_b,u_b}(u,v) = \frac{\sum (x(t)-\mean(x))(y(t)-\mean(y))}{(n-1)\sqrt{\var(x) \var(y)}},\label{PartialSpearmanEq}
\end{equation}
\noindent where sample means and variances in this equation
are computed using all the data but the sum is computed only
over the $t$ such that $x(t)+y(t) > 2l_b$ and $x(t)+y(t)<2u_b$. 
The partial Spearman correlation is the component of the Spearman
correlation of $u$ and $v$ which can be attributed to the points
for which the ranks, $x$ and $y$, lie between the boundary lines
we defined in
the unit square (Fig. \ref{PartialSpearmanFig}). 
The quantity $\cor_{0,0.5}(u,v)-\cor_{0.5,1}(u,v)$ 
is a way to measure asymmetry of tail associations between
$u$ and $v$. And this quantity is the same as 
$\cor_{0,0.5}(s_e,\tilde{f}_a)-\cor_{0.5,1}(s_e,\tilde{f}_a)$, 
because the partial Spearman correlation, like the Spearman
correlation itself, is based on ranks. Ranks are the same
for $u(t)$ and for $s_e(t)$ because these quantities are 
related by an increasing monotonic transformation (and likewise for 
$v(t)$ and $\tilde{f}_a(t)$). Thus, in the sense captured
by the statistic $\cor_{0,0.5}-\cor_{0.5,1}$, $s_e(t)$
and $\tilde{f}_a(t)$ have the same asymmetry of tail
association as the values $u(t)$ and $v(t)$ from which they
were constructed. Rank-based approaches to understanding the nature
of association between variables have been recommended, as such
approaches are not influenced by the marginal distributions of
the quantities (e.g., @Genest2007). Marginal distributions contain no information
about association between variables, as each marginal pertains to
only one of the variables.

<!--***DAN: Shyamolina, the below commented text makes me wonder if I am misunderstanding something, please
explain. The part that confuses me is the 2.5th and 97.5th quantile part, what is the meaning
of that and what does it have to do with extinction riks? I would have thought extinction risk
was just the fraction of 10000 sims that went extinct.

after each time step, we simulated model (\ref{eq_age_str}) for $25$ years
with $10000$ simulations. For each simulation, we started with initial population $(E,A)_{t=0} = (5,10)$
and following Eq. \ref{eq_age_str} we checked whether after each time step the total population $(E+A)_{t}$
was below a certain extinction threshold (we set it as $2$) or not. If total population was below that threshold then we 
considered population went extinct for that simulation. We also calculated $2.5^{th}$ and $97.5^{th}$ quantiles
of $10000$ simulations with the probability of computed extinction risk at each time step (see [`code`](https://github.com/sghosh89/ERC/blob/master/ext_risk_cop_age_str.R)).
-->

<!--density-dependence metapopulation model-->
To investigate Q2 from the \nameref{Introduction}, we considered spatialized, stochastic versions
of six density dependent population models, the Ricker, Hassell, Maynard Smith, Pennycuick, Verhulst, 
and Austin-Brewer models (Table \ref{tab_model_summary}).
All models took the general form $\vec{P} (t+1) = D \lambda(t) \vec{P}(t)$,
where $\vec{P}(t)$ is a vector of population densities, 
the $i^{\text{th}}$ component $P_i(t)$ representing
the population density in the $i^{\text{th}}$ habitat patch ($i=1,2,\ldots,n$) at time $t$. 
The $n \times n$ matrix $\lambda(t)$ was diagonal with $i^{\text{th}}$ diagonal entry 
$\lambda_i(t)$ depending on the model used (Table \ref{tab_model_summary}). The matrix $D$
was an $n \times n$ dispersal matrix. For simplicity, 
we assumed the $n$ habitat patches were arranged, evenly spaced, in a line.
We considered both "local" and "global" dispersal among patches in different simulations. 
For local dispersal, a fraction $d$ of 
each population dispersed during each time step, equally distributed to
the two or one nearest neighbors. For global dispersal the same fraction $d$ dispersed
equally to the other $n-1$ patches. When $d$ was $0$, dispersal did not occur and 
rescue effects were not possible.

@@

<!--***DAN: By the way, Shyamolina, these are not Hopf bifuractions - those cannot happen
with univariate models like these. These are called period-doubling bifucations.
If $r$ became too high and exceeded $r_{bf}$ - the parameter value responsible for Hopf-bifurcation -
model equilibrium $P_e$ became unstable. Beyond the Hopf-bifurcation point, the model showed sustained
oscillation and eventually led to chaos with increasing $r$. Appendix \ref{SM-model_stability}
summarized the stability analysis of each model in their deterministic form.
-->

We simulated our metapopulation models using environmental noises $\epsilon_i(t)$ (Table \ref{tab_model_summary}), 
independent through time, that were generated to have the same overall associations 
between all patches and in all simulations (Spearman's $\rho$ again `r srho_common`, up to sampling variation), 
but with different degrees and types of asymmetry of tail associations 
between patches in different simulations, as follows. For a simulation of length $T$,
we used methods that can generate $T$ independent $n$-tuples $(u_1(t),\ldots,u_n(t))$
for $t=1,\ldots,T$ with $\rho=$ `r srho_common` between $u_i(t)$ and $u_j(t)$ for all $i \neq j$
(up to sampling variation), but with either extreme left-, moderate left-, moderate right-,
or extreme right-tail association between all pairs of locations (as in Fig. \ref{fig_intro_cop}). 
Pairs $u_i(t)$ and $u_j(t)$ looked like Fig. \ref{fig_intro_cop} and were structured, statistically,
in the same way as the stochastic quantities $u(t)$ and $v(t)$ used for the non-spatial 
simulations described above. For any $i$, $u_i(t)$ was again uniformly distributed on 
the interval $(0,1)$. To obtain the values $\epsilon_i(t)$, the $u_i(t)$ were 
transformed so that the resulting numbers were
normally distributed with mean $0$ and standard deviation $\sigma$ (Table \ref{tab_model_summary}), 
for all $i$. This was 
accomplished with the inverse cummulative distribution function of the normal distribution
with mean $0$ and standard deviation $\sigma$. Because this was a monotonic transformation,
the Spearman correlation between $\epsilon_i(t)$ and $\epsilon_j(t)$ was the same 
as that between $u_i(t)$ and $u_j(t)$; likewise partial Spearman correlations were the same. 
The quantities $(u_1(t),\ldots,u_n(t))$
for $t=1,\ldots,T$ were generated as described in Appendix \ref{SM-makenoise}.

To calculate extinction risk for a given model and a given choice of tail association,
we simulated the model for $25$ time steps, $10000$ times, using $n=5$ habitat patches.
The initial condition $P_e$ was used in each habitat patch ($P_e$ is the equilibrium value of the 
one-patch, deterministic version of the model, Table \ref{tab_model_summary}).
After each time step, if $P_i(t)$ was less than $P_e/10$ then it was set to $0$.
The whole metapopulation was considered to have gone extinct if all patches were $0$
a time step $25$.
<!--***DAN: Shyamolina, please check the above two sentences, is this what you did?-->
Extinction risk was the fraction of 10000 metapopulations that went extinct.

@@

<!--**DAN: got to here 2019 10 29-->

<!--test hypothesis from Q2 : ext risk with density-dependence metapopulation model with r from different regime-->
To answer (Q2) of the \nameref{Introduction} for each metapopulation model,
we estimated extinction risk of each model using each form of tail 
association in the environmental noise,
across the range of $r$ indicated above that spanned under- and overcompensatory dynamics.
We chose equispaced values of $r$ from the 95\% quantiles 
of possible range ($r_{min},r_{bf}$), where $r_{min}$ was the minimum value of $r$ 
to get a positive, finite equilibrium $P_e$ (for details, see Appendix 
\ref{SM-model_stability}) to start with (at $t=0$). For each chosen $r$,
we calculated extinction risk after $25$ years with each metapopulation model 
where population could disperse locally (or globally) with varying dispersal 
rate $d= 0,0.1,0.2,..0.5$ and moderate to extreme left 
tail- and right-tail associated environmental noises with same Kendall's $\rho$ (= `r srho_common`)
in different simulations (for code see [`mydispersal.R`](https://github.com/sghosh89/ERC/blob/master/mydispersal.R)).

We plotted extinction risks along the y-axis
for both left tail and right tail associated noises 
against the varying range of $r$ along x-axis when no dispersal ($d=0$) was considered.
We excluded the possibility of dispersal to avoid any rescue effect among patches. 
For each model with extreme (or moderate) left- and right-tail associated noises, we calculated 
2.5$^{th}$ and 97.5$^{th}$ quantiles
of 10000 simulations with the probability of computed
extinction risks at each $r$. We compared $r_c$ and the value of $r$ at which 
95\% CI of extinction risks from both noises overlapped
with each other. This could give us an idea of how extinction risks affected by the model parameter 
$r$ that controlled the underlying dynamics of the model, i.e., undercompensatory 
or overcompensatory.

@@

<!--Q1 : age-structured results : linear and nonlinear model with asymmetric copula-->

For single patch linear age-structured models, we found higher extinction risk for moderate left-tail 
association (modeled by Clayton copula) in fluctuating vital 
rates compared with the one when vital rates were right tail dependent (modeled by Survival Clayton copula).
It was expected because an extreme "bad" year (e.g. with severe cold) can suppress 
the fecundity of adult and survival/ maturation of eggs simultaneously 
and thereby can increase the extinction risk of the population. For non-linear age-structured 
models, we got a similar trend for extinction risks with the different tail associations (left and right).
The non-linear model showed higher extinction risk than its corresponding 
linear analog as it was evident from Eq. \ref{eq_vitalrates} with finite $\Omega = 100$. 
Fig. \ref{fig_extrisk_agestr} represents this scenario when vital rates have moderate to extreme tail association in either tail but with same Kendall's correlation ($\rho =$ `r srho_common`). Extinction risks for two different tail-dependent vital rates were not strinkingly different (e.g. see pair of red and
blue lines in Fig. \ref{fig_extrisk_agestr}A or B) and we anticipated the difference being partly 
dependent on the degree of tail association. Therefore we repeated the same analysis with 
vital rates coming from extreme-tail dependent (left and right in two different simulations)
copulas and got similar results but with much prominent difference between their extinction risks
(Fig. \ref{fig_extrisk_agestr}C or D). This result answered (Q1) and showed extinction risk of 
a population could be considered as a direct manifestation of tail association property 
between vital rates of different age-class
of that population. <!--discussion later: add Tuljapurkar's approximation-->

@@

 <!--FIG : age structure model results : linear, nonlinear-->
\begin{figure*}[!h]
\textbf{ \hspace{5.4 cm} (A) \hspace{4 cm} (B)} \\
\vspace{-0.9 cm}
\begin{center}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_C_SC_linear.pdf}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_C_SC_nonlinear_omg_100.pdf}\\
\end{center}
\textbf{ \hspace{5.4 cm} (C) \hspace{4 cm} (D)} \\
\vspace{-0.9 cm}
\begin{center}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_extremetail_linear.pdf}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_extremetail_nonlinear_omg_100.pdf}
\caption{Extinction risk against time from single patch age-structured linear ($\Omega = \infty$) $(A, C)$ and nonlinear ($\Omega = 100$) $(B, D)$ models for $10000$ simulations; shading represents 95\% confidence interval. $(A, B)$ show extinction risk with time when fecundity of adults and survival of eggs have a moderate left tail (LT) or a moderate right tail (RT) association, $(C, D)$ show the analogous scenario but with the extreme tail association between vital rates. \label{fig_extrisk_agestr}}
\end{center}
\end{figure*}

@@

<!--Summary tables for 6 nonlinear models used-->
```{r read_tab_model_summary,echo=F, results='asis',message=F}
library(kableExtra)
library(dplyr)
dt<-readRDS("./Results/tab_model_summary.RDS")
knitr::kable(dt, escape=F,
             format="latex",digits=3, align="c",table.env='table*',
             caption = "Summary of the six density-dependent stochastic metapopulation models we used. The population growth rate in habitat patch $i$ at time $t$, $\\lambda_i(t)$ (second column), is $P_i(t+1)/P_i(t)$, where $P_i(t)$ is the population in patch $i$ as time $t$. The growth rate depends on model parameters, $P_i(t)$, and the environment $\\epsilon_i(t)$ in patch $i$ at time $t$. $\\epsilon_i(t)$ was normally distributed with mean $0$ and standard deviation $\\sigma$, and values were generated independently across times. $P_e$ is the equilibrium of the deterministic, one-patch version of the model. $r_c$ is the value of $r$ for which a transition occurs from under- to overcompensatory dynamics (see main text). $r_{bf}$ is the value of $r$ for which a transition occurs from a stable to an unstable equilbrium (a bifurcation). We use only $r<r_{bf}$. Dispersal between patches was also implemented as described in the main text.\\label{tab_model_summary}",
             booktabs=T, linesep = "\\addlinespace"
             )%>%column_spec(c(1,3,5,6),width="1.4 cm")%>%kable_styling(latex_options = "basic",position = "center")%>%column_spec(2, width = "3.8 cm")
```
<!--***DAN: Shyamolina, can you please make the header of the second column read simply \lambda_i(t)?-->

>>>>

<!--FIG : pedagogical fig showing 4 types of tail associations: 
extreme left, moderate left, moderate right, extreme right-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=15 cm]{./Results/introfig_copula.pdf}
\caption{Two stochastic quantities can have same Spearman 
correlation, but different degrees of tail association: 
(A) extreme left-tail association, (B) moderate left-tail association, (C) 
moderate right-tail association, and (D) extreme right-tail association.
In all cases, Spearman's $\rho$ was the same, $\rho = 0.875$,
up to sampling variance (the sample correlation for each dataset is displayed
on its panel). The extreme left-tail (respectively, right-tail) dependent 
case is perfectly correlated below (respectively, above) the threshold of 0.
See Methods for information on how bivariate noise was generated with
asymmetric tail associations.
\label{fig_intro_cop}}
\end{center}
\end{figure*}
<!--***DAN: Shyamolina, you put a negative sign in front of the Spearman correlations
on the C and D panels, but that is erroneous. The Spearman correlations are positive
in all four cases! Please fix!-->

@@

The stochastic quantities $s_e(t)$ and $\tilde{f}_a(t)$ were generated 
to have the same overall association (Spearman's $\rho$ equal to $0.875$,
up to sampling variance), but different degrees and types of 
asymmetry of tail association in different simulations, as follows.
For a simulation of length $T$, we used methods that can 
generate $T$ independent pairs $(u(t),v(t))$ for $t=1,\ldots,T$ with
$\rho=0.875$ (up to sampling variance), but with either extreme left-,
moderate left-, moderate right-, or extreme right-tail association 
(e.g., Fig. \ref{fig_intro_cop}). The methods we used generated
$(u(t),v(t))$ such that $u(t)$ was uniformly distributed on $(0,1)$,
and likewise for $v(t)$. We then applied a monotonic transformation
to the $u(t)$ so that the resulting numbers were distributed according to
a beta distribution with parameters
$\alpha=3$ and $\beta=2$, 
and we used those values for the $s_e(t)$. We also applied a (different)
monotonic transformation to the $v(t)$ so that the resulting numbers were
distributed according to a gamma distribution with parameters $k=2$
and $\theta=1$, and we used those values for the $\tilde{f}_a(t)$. Because
the $u(t)$ and $v(t)$ were uniformly distributed, the inverse 
cummulative distribution functions of the desired beta and
gamma distributions were the transformations. Because these
were monotonic transformations, the Spearman correlation of 
$s_e(t)$ and $\tilde{f}_a(t)$ was the same as that of 
$u(t)$ and $v(t)$. The quantities $(u(t),v(t))$ for $t=1,\ldots,T$ were
generated as described in Appendix \ref{SM-makenoise}.

<!--calculation of ext risk with age-structured model-->
To calculate extinction risk for a given $\Omega$ and choice of tail dependence, 
we simulated the model for $25$ time steps, $10000$ times,
each simulation starting from $E(t)=5$ and $A(t)=10$. The population for a simulation
was considered to have gone extinct if the total population $E(t)+A(t)$ ever 
went below an extinction threshold (we used $2$) within the 
$25$-time-step time horizon. Extinction risk was the fraction of $10000$
simulations that went extinct. Confidence intervals (95\%) of extinction
risk were also computed based on a binomial distribution.

The variables $u(t)$ and $v(t)$ are visibly asymmetrically tail associated (Fig. \ref{fig_intro_cop}),
but we have instead used the transformed variables $s_e(t)$ and $\tilde{f}_a(t)$
in our model. In what sense are $s_e(t)$ and $\tilde{f}_a(t)$ asymmetrically
tail associated? Given two boundary quantities $l_b$ and $u_b$ such
that $0 \leq l_b < u_b \leq 1$, and letting $x(t)$ be the rank of 
$u(t)$ in the set $\{u(t) | t=1,\ldots,T\}$ and letting 
$y(t)$ be the rank of 
$v(t)$ in the set $\{v(t) | i=1,\ldots,T\}$ (smallest elements
are here defined to have rank 1),
consider the portions of the 
lines $x+y = 2l_b$ and $x+y=2u_b$ which lie within the unit square
(Fig. \ref{PartialSpearmanFig}). 
<!--***DAN: Shyamolina, please make this figure. You can start from the code that 
make the similar figure in BIVAN (Fig. 7), but change it slightly so the figure is not
*exactly* the same. E.g., you could just switch from Clayton to survival Clayton data,
and maybe slightly change the locations of the bands. The caption can also be similar,
but obviously should not mention D^2 or P, and wording should be changed slightly.
The caption will also have to say something like "This figure was reproduced largely 
following Fig. 7 of Ghosh et al".-->
We can then define the *partial Spearman correlation*,
\begin{equation}
\cor_{l_b,u_b}(u,v) = \frac{\sum (x(t)-\mean(x))(y(t)-\mean(y))}{(n-1)\sqrt{\var(x) \var(y)}},\label{PartialSpearmanEq}
\end{equation}
\noindent where sample means and variances in this equation
are computed using all the data but the sum is computed only
over the $t$ such that $x(t)+y(t) > 2l_b$ and $x(t)+y(t)<2u_b$. 
The partial Spearman correlation is the component of the Spearman
correlation of $u$ and $v$ which can be attributed to the points
for which the ranks, $x$ and $y$, lie between the boundary lines
we defined in
the unit square (Fig. \ref{PartialSpearmanFig}). 
The quantity $\cor_{0,0.5}(u,v)-\cor_{0.5,1}(u,v)$ 
is a way to measure asymmetry of tail associations between
$u$ and $v$. And this quantity is the same as 
$\cor_{0,0.5}(s_e,\tilde{f}_a)-\cor_{0.5,1}(s_e,\tilde{f}_a)$, 
because the partial Spearman correlation, like the Spearman
correlation itself, is based on ranks. Ranks are the same
for $u(t)$ and for $s_e(t)$ because these quantities are 
related by an increasing monotonic transformation; and likewise for 
$v(t)$ and $\tilde{f}_a(t)$. Thus, in the sense captured
by the statistic $\cor_{0,0.5}-\cor_{0.5,1}$, $s_e(t)$
and $\tilde{f}_a(t)$ have the same asymmetry of tail
association as the values $u(t)$ and $v(t)$ from which they
were constructed. Rank-based approaches to understanding the nature
of association between variables have been recommended, as such
approaches are not influenced by the marginal distributions of
the quantities (e.g., @Genest2007). Marginal distributions contain no information
about association between variables, as each marginal pertains to
only one of the variables.

<!--density-dependence metapopulation model-->
To investigate Q2 from the \nameref{Introduction}, we considered spatialized, stochastic versions
of six density dependent population models, the Ricker, Hassell, Maynard Smith, Pennycuick, Verhulst, 
and Austin-Brewer models.
All models took the general form $\vec{P} (t+1) = D \lambda(t) \vec{P}(t)$,
where $\vec{P}(t)$ is a vector of population densities, 
the $i^{\text{th}}$ component $P_i(t)$ representing
the population density in the $i^{\text{th}}$ habitat patch ($i=1,2,\ldots,n$) at time $t$. 
The $n \times n$ matrix $\lambda(t)$ was diagonal with $i^{\text{th}}$ diagonal entry 
$\lambda_i(t)$ depending on the model used (Table \ref{tab_model_summary}). The matrix $D$
was an $n \times n$ dispersal matrix. For simplicity, 
we assumed the $n$ habitat patches were arranged, evenly spaced, in a line.
We considered both "local" and "global" dispersal among patches in different simulations. 
For local dispersal, a fraction $d$ of 
each population dispersed during each time step, equally distributed to
the two or one nearest neighbors. For global dispersal the same fraction $d$ dispersed
equally to the other $n-1$ patches. When $d$ was $0$, dispersal did not occur and 
rescue effects were not possible.

@@

<!--***DAN: By the way, Shyamolina, these are not Hopf bifuractions - those cannot happen
with univariate models like these. These are called period-doubling bifucations.

If $r$ became too high and exceeded $r_{bf}$ - the parameter value responsible for Hopf-bifurcation -
model equilibrium $P_e$ became unstable. Beyond the Hopf-bifurcation point, the model showed sustained
oscillation and eventually led to chaos with increasing $r$. Appendix \ref{SM-model_stability}
summarized the stability analysis of each model in their deterministic form.
-->

We simulated our metapopulation models using environmental noises $\epsilon_i(t)$ (Table \ref{tab_model_summary}), 
independent through time, that were generated to have the same overall associations 
between all patches and in all simulations (Spearman's $\rho$ again `r srho_common`, up to sampling variance), 
but with different degrees and types of asymmetry of tail associations 
between patches in different simulations, as follows. For a simulation of length $T$,
we used methods that can generate $T$ independent $n$-tuples $(u_1(t),\ldots,u_n(t))$
for $t=1,\ldots,T$ with $\rho=$ `r srho_common` between $u_i(t)$ and $u_j(t)$ for all $i \neq j$
(up to sampling variance), but with either extreme left-, moderate left-, moderate right-,
or extreme right-tail association between all pairs of locations (as in Fig. \ref{fig_intro_cop}). 
Pairs $u_i(t)$ and $u_j(t)$ looked like Fig. \ref{fig_intro_cop} and were structured, statistically,
in the same way as the stochastic quantities $u(t)$ and $v(t)$ used for the non-spatial 
simulations described above. For any $i$, $u_i(t)$ was again uniformly distributed on 
the interval $(0,1)$. To obtain the values $\epsilon_i(t)$, the $u_i(t)$ were 
transformed so that the resulting numbers were
normally distributed with mean $0$ and standard deviation $\sigma$ (Table \ref{tab_model_summary}), 
for all $i$. This was 
accomplished with the inverse cummulative distribution function of the normal distribution
with mean $0$ and standard deviation $\sigma$. Because this was a monotonic transformation,
the Spearman correlation between $\epsilon_i(t)$ and $\epsilon_j(t)$ was the same 
as that between $u_i(t)$ and $u_j(t)$; likewise partial Spearman correlations were the same. 
The quantities $(u_1(t),\ldots,u_n(t))$
for $t=1,\ldots,T$ were generated as described in Appendix \ref{SM-makenoise}.

To calculate extinction risk for a given metapopulation model, given parameters,
and a given choice of tail association,
we simulated the model for $25$ time steps, $10000$ times, using $n=5$ habitat patches.
The equilibrium value of the 
one-patch, deterministic version of the model, $P_e$ (Table \ref{tab_model_summary})
was used as the initial condition in all patches and simulations.
After each time step, if $P_i(t)$ was less than $P_e/10$ then it was set to $0$.
The whole metapopulation was considered to have gone extinct if all patches were $0$
a time step $25$.
<!--***DAN: Shyamolina, please check the above two sentences, is this what you did?-->
Extinction risk was the fraction of 10000 metapopulations that went extinct. 
Estimates of extinction risk were plotted together with 95\% confidence intervals
based on a binomial distribution.

@@

<!--test hypothesis from Q2 : ext risk with density-dependence metapopulation model with r from different regime-->
To answer (Q2) of the \nameref{Introduction}, 
we estimated extinction risk for each metapopulation model, using each form of tail 
association for the environmental noise,
using values of $r$ that included under- and overcompensatory dynamics,
as indicated above. We chose values of $r$ within a range, the
range specified as follows. The second part of Q2 (about how density dependence mediates the
influence of asymmetric tail associations on extinction risk) was addressed by comparing
results for under- and overcompensatory model regimes.
Defining $r_{min}$ as the minimum value of $r$ 
for which $P_e$ was non-negative; and defining
$r_{bf}$ as the value of $r$ at which a bifurcation occurs, the range of
$r$ used spanned 95\% of the interval ($r_{min},r_{bf}$). For details, see Appendix 
\ref{SM-model_stability}. The range of $r$ encompassed $r_c$, so included 
a transition from under- to overcompensatory dynamics. For each value of $r$
used within the range, we calculated extinction risk as indicated above.
Dispersal parameter values $d= 0,0.1,0.2,..0.5$ were also used in different 
simulations. We plotted extinction risk as a function of $d$ for two
fixed values of $r$, one producing under- and one overcompensatory dynamics.
For $d=0$ we also plotted extinction risk as a function of $r$. All code for 
this project is available at https://github.com/sghosh89/ERC.

<!--
We plotted extinction risks along the y-axis
for both left tail and right tail associated noises 
against the varying range of $r$ along x-axis when no dispersal ($d=0$) was considered.
We excluded the possibility of dispersal to avoid any rescue effect among patches. 
For each model with extreme (or moderate) left- and right-tail associated noises, we calculated 
2.5$^{th}$ and 97.5$^{th}$ quantiles
of 10000 simulations with the probability of computed
extinction risks at each $r$. We compared $r_c$ and the value of $r$ at which 
95\% CI of extinction risks from both noises overlapped
with each other. This could give us an idea of how extinction risks affected by the model parameter 
$r$ that controlled the underlying dynamics of the model, i.e., undercompensatory 
or overcompensatory.
-->

@@

<!--Q1 : age-structured model results-->
For our age-structured, non-spatial model, extinction risks were higher when 
associations between the vital rates $s_e$ and $f_a$ were principally for low values
of these vital rates (left-tail association), and were lower when associations
were principally for higher values; but the effect was minor for this model (Fig. \ref{fig_extrisk_agestr}).
Though extinction risks were overall higher for the density-dependent version of our model compared
to density-independent version, extinction risk differences across tail association 
scenarios were similar for the two versions. Because the same Spearman correlation (up to sampling
variance) was used in all simulations, the differences between lines in the panels
of Fig. \ref{fig_extrisk_agestr} are attributable solely to the changes we made in the
asymmetry of tail associations, and answer Q1. Though differences were significant (confidence
bands in Fig. \ref{fig_extrisk_agestr}, which are very narrow, did not overlap),
they were also small, corresponding to only a few percentage points of additional extinction
risk after 25 years. A potential reason for this outcome is elaborated in the Discussion.
<!--discussion later: add Tuljapurkar's approximation stuff-->

@@

 <!--FIG : age structure model results : linear, nonlinear-->
\begin{figure*}[!h]
\textbf{ \hspace{5.4 cm} (A) \hspace{4 cm} (B)} \\
\vspace{-0.9 cm}
\begin{center}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_C_SC_linear.pdf}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_C_SC_nonlinear_omg_100.pdf}\\
\end{center}
\textbf{ \hspace{5.4 cm} (C) \hspace{4 cm} (D)} \\
\vspace{-0.9 cm}
\begin{center}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_extremetail_linear.pdf}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_extremetail_nonlinear_omg_100.pdf}
\caption{Extinction risk versus time for our stochastic, age-structured model; risks
were higher when associations between vital rates were stronger in the left compared to
the right tails of the rates, but only slightly so. Results are for the linear
version of the model ($\Omega = \infty$; A, C) and for the nonlinear model with density
dependence in the fecundity rate ($\Omega = 100$; B, D). Results show cases of moderate
(A, B) and extreme (C, D) asymmetry of tail association. The solid line is the estimated extinction risk and the area of shading, which is barely visible because it is so narrow, show 95\% confidence intervals for extinction risk.\label{fig_extrisk_agestr}}
\end{center}
\end{figure*}

@@

<!--Summary tables for 6 nonlinear models used-->
```{r read_tab_model_summary,echo=F, results='asis',message=F}
library(kableExtra)
library(dplyr)
dt<-readRDS("./Results/tab_model_summary.RDS")
knitr::kable(dt, escape=F,
             format="latex",digits=3, align="c",table.env='table*',
             caption = "Summary of the six density-dependent stochastic metapopulation models we used. The population growth rate in habitat patch $i$ at time $t$, $\\lambda_i(t)$ (second column), is $P_i(t+1)/P_i(t)$, and defines how the model works within each patch. Here $P_i(t)$ is the population in patch $i$ at time $t$. The growth rate depends on model parameters, $P_i(t)$, and the environment $\\epsilon_i(t)$ in patch $i$ at time $t$. $\\epsilon_i(t)$ was normally distributed with mean $0$ and standard deviation $\\sigma$, and values were generated independently across times. $P_e$ is the equilibrium of the deterministic, one-patch version of the model. $r_{min}$ is the smallest $r$, given fixed values of the other parameters, for which $P_e$ was non-negative. $r_c$ is the value of $r$ for which a transition occurs from under- to overcompensatory dynamics (see main text). $r_{bf}$ is the value of $r$ for which a transition occurs from a stable to an unstable equilbrium (a bifurcation; see main text). We used $r$ such that $r_{min}<r<r_{bf}$. Dispersal between patches was also implemented (see main text).\\label{tab_model_summary}",
             booktabs=T, linesep = "\\addlinespace"
             )%>%column_spec(c(1,3,5,6),width="1.4 cm")%>%kable_styling(latex_options = "basic",position = "center")%>%column_spec(2, width = "3.8 cm")
```
<!--***DAN: Shyamolina, can you please make the header of the second column read simply \lambda_i(t)? Also, can you add a column for r_min, please?-->
